% !TEX root = diss.tex

\chapter{Acknowledgements}      % Optional

\begin{doublespace}

I would like to first express my deepest gratitude to my supervisor, Prof. Gino
DiLabio, for his guidance throughout the course of my degree. His depth of
knowledge and continuing enthusiasm were always there to support and inspire
me. I am truly indebted to him and cannot say thank you enough. I would also
like to thank all the past and current members of the DiLabio group, without
whom the computational research path would have been exceedingly isolated. A
special thank you to Dr. Alberto Otero de la Roza for all his help with many
research and technical matters.

Thank you to Dr. Steve McNeil, Dr. Fred Menard, Dr. Paul Shipley, Dr. Jim
Bailey, and Dr. Sandra Mecklenburg for their advice, knowledge, and dedication
to teaching. Your collective efforts as instructors, mentors, and advisors have
helped tremendously.

I would also like to thank my family for their endless support, encouragement,
and love: Michael, Mama, Papa, Nicola, Rob, and Arlene. Words cannot express how
much you all mean to me.

And finally I would like to thank Grace. Your never ending support and love have
been fundamental to my success.

\end{doublespace}

% \chapter{Dedication} % Optional
% The dedication is usually quite short, and is a personal rather than
% an academic recognition.  The \emph{Dedication} does not have to be
% titled, but it must appear in the table of contents.  If you want to
% skip the chapter title but still enter it into the Table of Contents,
% use this command \verb|\chapter[Dedication]{}|.
